---
title: "Predicting Post Injury Play Time For NFL Players"
author: 'Group D: Arthur, Donna, Kaitlyn, Michael'
date: "5/7/2024"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  word_document: default
---

```{r include=FALSE}
library(mosaic)
library(GGally)
require(tidyr)
require(car)
library(broom)
library(tidyverse)
require(lattice)
library(gridExtra)
library(grid)
library(leaps)
library(lubridate) # finding age at a given date
options(scipen=5)
```

```{r include=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)
trellis.par.set(theme=theme.mosaic())  
options(digits=5)
```


```{r, include=FALSE, warning = FALSE, echo = FALSE}
###############################################################################################
#getwd() in console to find the directory where the csv file is stored on your computer
#For your computer you need to change the path so you can reed the file
#alldata <- read.csv("/home/class26/FIRSTPARTOFEMAIL/ConcussionInjuries2012-2014.csv") 

##Donna's computer
#df1 <- read.csv("C:/Users/donna/OneDrive/Documents/STAT230/Stats230FinalProject/ConcussionInjuries2012-2014.csv")
#df2 <- read.csv("C:/Users/donna/OneDrive/Documents/STAT230/Stats230FinalProject/HeadInjuredPlayers.csv")


#For Kaitlyn's computer
df1 <- read.csv("/home/class27/khuang27/Stats230FP/ConcussionInjuries2012-2014.csv")
df2 <- read.csv("/home/class27/khuang27/Stats230FP/HeadInjuredPlayers.csv")

#For Arthur's computer
#df1 <- read.csv("/home/class26/anwobi26/Stats230_proj/ConcussionInjuries2012-2014.csv")
#df2 <- read.csv("/home/class26/anwobi26/Stats230_proj/HeadInjuredPlayers.csv")

###############################################################################################
##Merge
## gets rid of players who had multiple injuries in the three years
mergeData <- merge(x = df2, y = df1[!duplicated(df1$Player), ], by = "Player", all.x = TRUE)
mergeData <- filter(mergeData, Reported.Injury.Type == "Concussion") ## filter so there's only concussions 

## qual --> quant var
mergeData$Play.Time.After.Injury = parse_number(mergeData$Play.Time.After.Injury)
mergeData$Average.Playtime.Before.Injury = parse_number(mergeData$Average.Playtime.Before.Injury)
###############################################################################################
##rename variables
nameData <- rename(mergeData, OppTeam = Opposing.Team, PostIPT = Play.Time.After.Injury, PreseasonInjury = Pre.Season.Injury., OnWinningTeam = Winning.Team., amtTimeInjured = Weeks.Injured, WeekOfInjury = Week.of.Injury, UnknownInjury = Unknown.Injury., ReportedTypeOfInjury = Reported.Injury.Type, PreIPT = Average.Playtime.Before.Injury, TotalSnaps = Total.Snaps, amtGamesMissed = Games.Missed, GM12 = X2012.2013...Games.Missed, GM13 = X2013.2014...Games.Missed, GM14 = X2014.2015...Games.Missed, I12 = X2012.2013...Number.of.Injuries , I13 = X2013.2014...Number.of.Injuries, I14 = X2014.2015...Number.of.Injuries, TotalInjuries = Total.Number.of.Injuries..2012.2014., PosDuringInjury = Roles.during.injuries, DOB = Date.of.Birth, AgeAtConcussion = Age.first.concussion..2012.2014., totalGamesMissed = Total.Games.Missed..2012.2014., TeamDurConcussion = Team.s..during.concussion.incidents.2012.2014, DOI = Date)

###############################################################################################
## select variables 
studyD <- nameData %>% 
  select(PostIPT, PreseasonInjury, Season, amtGamesMissed, PreIPT, WeekOfInjury, TotalSnaps, Player, DOB, DOI, AgeAtConcussion, Position, Team, OppTeam, OnWinningTeam)

## getting the current age 
# Convert DOB to date objects
studyD$DOB <- as.Date(as.character(studyD$DOB), format = "%d/%m/%Y")

# Calculate current age
given_date <- as.Date("2024-04-01")
current_age_in_years <- as.numeric(interval(studyD$DOB, given_date) / dyears(2))
studyD$CurrentAge <- round(current_age_in_years, 1)

#############################################################################
## this re-categorizes position to offense and defense
studydata <- studyD %>% 
    mutate(PositionC = case_when(
    Position %in% c("Defensive End", "Defensive Tackle", "Safety", "Comerback" ) ~ "defense",
    Position %in% c("Long Snapper", "Wide Receiver", "Tight End", "Linebacker", "Quarterback", "Center", "Guard", "Full Back", "Running Back", "Offensive Tackle") ~ "offense"))
#############################################################################


#############################################################################
## showing why we are removing preseason injury. 
tally(~PreseasonInjury,data = studydata)

m6a <- gf_bar(~PreseasonInjury,data = studydata)%>%
  gf_labs(title = "Fig. 6a. Pre-season Injury", x = "Pre-season Injury (Yes/No)") 

## filters out players who have gotten preseason injury bc not enough observations. 
##population
finalData <- filter(studydata, PreseasonInjury == "No") 

#############################################################################
## data for the season we are predicting
predictData <-filter(studydata, Season == "2014/2015") 

##sample
## selecting years
studydata <- filter(studydata, Season != "2014/2015") ## our sample data
finalData1 <- select(studydata, "PostIPT" ,"PreIPT", "amtGamesMissed", "WeekOfInjury", "AgeAtConcussion", "PreseasonInjury", "Position","Team", "OnWinningTeam", "PositionC", "Player")

# mutates amt games missed to categorical variable. 
finalData1 <- mutate(finalData1, amtGamesMissedC = cut(amtGamesMissed, breaks = c(-1, 0.5, 1.5, 20), labels = c("no games", "one game", "two or more games"), include.lowest = TRUE))

finalData1 <- filter(finalData1, amtGamesMissedC %in% c("no games", "one game", "two or more games"))
```


# Abstract
For our project, we looked at different factors that could influence post-injury playtime for players in the NFL who experienced concussions. Specifically we were interested in what factors maximized or minimized post-injury play time to evaluate potential adjustments to the NFL season structure, and to help players understand what factors maximize their playtime post-injury. We examined factors such as position, amount of games missed, age at concussion, preseason injury (Yes/No), week of injury, and finally pre-injury playtime. Using data from the 2012-2014 NFL seasons and running analysis on each of these factors to predict post-injury playtime, we found that pre-injury playtime and amount of games missed were the only significant predictors. These predictors were able to account for approximately 51.5% of the variability. With a more robust data set that includes player data for at least 5-10 NFL seasons, we believe we can find stronger relationships and satisfy conditions to find more significant predictors.


# Background and Meaning
Concussions have been one of the most common injuries in the NFL for some time now and due to their potential long term effects, how can the NFL help players return from injury and still be able to play at the same level before their injury? In 2018, after data from the previous season showed an increase in diagnosed concussions in players, the NFL Adjusted its concussion protocol to better protect the health and safety of its players. In the past there have also been various rule changes that affect tackling technique and currently the NFL is in discussion to decide to shorten the amount of preseason games or to get rid of the preseason completely. Our findings could potentially provide evidence for additional alterations to the NFL season structure, and also test if past changes have had an effect on maximizing the playtime of players post injury. Our null hypothesis for this study was that there is no relationship between Post-Injury Play Time and the variables we selected to analyze. Our alternative hypothesis was that there is at least one significant relationship between Post-Injury Play Time and the variables we selected. 

# **Methods**

## Data Set
The Concussions in the NFL (2012-2014) data we used contains the data set of NFL players who were recorded for sustained concussions, head injuries, or illnesses during the season from the years 2012 to 2014. This data set details a number of factors surrounding these players and their injuries such as pre-season injury history, position, average play time after injury, the number of weeks injured, the season they got injured, and more. The population of this data set that was sampled is NFL players during the 2012-2014 seasons that obtained a concussion. We filtered the data set for only concussed players to determine the impact of concussions on players’ total playing time.

## Variables

### Response Variable:

**Post-Injury Play Time:** The variable post-injury playtime (PostIPT) is our response variable and it represents the average playtime for a player after their concussion injury, measured by downs played. 

### Explanatory Variables:
The variables pre-injury playtime, week of injury, age of concussion, amount of games missed, position, and preseason injury were the explanatory variables.

**Pre-injury playtime:** Pre-injury playtime (PreIPT) is a quantitative variable that represents the average playtime for a player pre-concussion injury measured by downs played. There was one unusual point, Chris Prosinski (PreIPT = 0 downs, PostIPT = 82 downs) who we removed for our bivariate analysis.  

**Week of injury:** Week of injury (WeekOfInjury) is a quantitative variable that represents the week during the NFL season of 17 weeks that a player sustained a concussion. 

**Age of concussion:** Age of concussion (AgeOfConcussion) is a quantitative variable that represents the player’s age in years at which they sustained the injury during the selected seasons.  

**Amount of games missed:** The amount of games missed (amtGamesMissed) variable represents the amount of games in a season that a player misses due to a concussion. The distribution is strongly skewed to the right, with the majority of players not missing any games. In order to make the heavily skewed distribution usable for our analysis, we decided to transform this variable into a categorical variable. The categories within this transformed variable include “no games”, “one game”, and “two or more games”. Majority of the players missed no games.

**Position:** Position (Position) is a qualitative variable that represents the player’s position. The categories of this variable include: defensive end (1), defensive tackle (2), safety (3), comerback (4), long snapper (5), wide receiver (6), tight end (7), linebacker (8), quarterback (9), center (10), guard (11), full back (12), running back (13) and offensive tackle (14). To make this variable easier for further analysis, we transformed this variable into two broader categories: "Offense" and "Defense".

**Preseason injury:** Preseason injury (PreseasonInjury) is a qualitative variable that represents if a player got injured during the pre-season of their respective season of their concussion injury. Categories are “yes”, indicating a preseason injury, and “no”, indicating no injuries sustained during the pre-season. Majority of the players in the sample did not sustain a preseason injury so we removed this variable for our analysis.

## Statistical Methods

  We constructed box plots to relate Post-Injury Play Time to our qualitative variables: Position, and Amount of Games Missed. We constructed scatter plots to relate Post-Injury Play Time to our quantitative variables: Pre-Injury Play Time, Week of Injury, and Age at Concussion. We checked conditions using residuals plots for equal variance and linearity and qq-plots for normality. We obtained one final model with Pre-Injury Play Time as a predictor. We evaluated the significance of this simple linear regression(SLR) model using the p-value for the F-statistic. We then removed unusual points using a combined leverage vs. standard residuals plot and Cooks Distance plot. We only removed one observation for the sake of preserving data. 
  For multiple linear regression we obtained a Variance Inflation Factor(VIF) for each predictor to detect issues with multicollinearity. Using p-values for the t-test, we reduced our kitchen sink model to our second final model with Pre-injury Play Time and Games Missed as a categorical variable. We checked conditions using a normal QQ-plot (for normality of residuals) and residuals vs. fitted plot (for equal variance and linearity). We evaluated the significance of this model using the p-value for F-statistic. Since our simple linear regression model was nested in our MLR model, we conducted a nested F-test and used the p-value associated with the F-statistic to evaluate if the MLR model was better overall. 

# **Results**

## Univariate Exploratory Data Analysis

### Response Variable

Our response variable for this study is Post-Injury Play Time.  

```{r, warning = FALSE, echo = FALSE, fig.height = 2, fig.width = 3}
gf_dhistogram(~PostIPT, data = finalData1)%>%
  gf_dens()%>%
  gf_labs(title = "Fig. 1. Post-Injury Play Time")
```
```{r, warning = FALSE, echo = FALSE, results='hide'}
favstats(~PostIPT, data = finalData1)
```
**Fig. 1.** The distribution is bimodal, with modes at around 20 downs and 62 downs, and symmetric. The median is 49 downs and the IQR is 42.5 downs. The mean is 47.37 downs and the standard deviation is 24.64 downs.

### Explanatory Variables - Qualitative

Our qualitative explanatory variables are Amount of Games Missed, Position, and Pre-Season Injury.

```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 10}
AmtGamesMissedUni <- gf_bar(~amtGamesMissedC, data = finalData1, fill = c("darkred", "darkblue", "darkgreen")) %>%
  gf_dens() %>%
  gf_labs(title = "Amount of Games Missed (Transformed)",  x = "Amount of Games Missed")

AmtGamesMissedUniQuant <- gf_dhistogram(~amtGamesMissed, data = finalData1)%>%
  gf_dens()%>%
  gf_labs(title = "Amount of Games Missed (Non-Transformed)", x = "Amount of Games Missed")

PositionUni <- gf_bar(~ PositionC, data = finalData1, fill = c("pink", "skyblue"))%>%
  gf_labs(title = "Position", x = "Position")

PreSeasonInjuryUni <- gf_bar(~ PreseasonInjury, data = studydata, fill = c("pink", "darkgreen"))%>%
  gf_labs(title = "Pre-Season Injury", x = "Pre-Season Injury")

grid.arrange(AmtGamesMissedUni, AmtGamesMissedUniQuant, ncol = 2)

grid.arrange(PositionUni, PreSeasonInjuryUni, ncol = 2)
```
```{r, warning = FALSE, echo = FALSE, results='hide'}
favstats(PostIPT ~ amtGamesMissedC, data = finalData1)
favstats(PostIPT ~ PositionC, data = finalData1)
tally(~PreseasonInjury,data = studydata)
```
**Fig. 2** Amount of Games Missed, originally quantitative, is heavily right-skewed, indicating that the majority of players did not miss multiple games. We resultantly transformed it into a categorical variable with the categories "Less than 1 games missed" (54.79%), "One game missed" (33.79%), and "Two or more games missed" (11.42%). Regarding position, the categories and corresponding distribution are: defensive end (5.31%), defensive tackle (3.38%), safety (13.04%), comerback (14.98%), long snapper (0.48%), wide receiver (14.98%), tight end (9.18%), linebacker (10.14%), quarterback (4.35%), center (1.93%), guard (6.76%), full back (0.97%), running back (10.63%) and offensive tackle (3.87%). We transformed this variable into two broader categories: "Offense" (59.82%) and "Defense" (40.18%). Finally, regarding Pre-Season Injury, there was an imbalance of observations; only 9 players sustained a pre-season injury in the first two seasons, while 207 players did not. Thus, we removed this variable for our bivariate analysis. 


### Explanatory Variables - Quantitative

Our quantitative variables are Pre-Injury Play-Time, Week of Injury and Age at Concussion. 

```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 15}
PreIPTUni <- gf_dhistogram(~PreIPT, data = finalData1) %>%
  gf_dens()%>%
  gf_labs(title = "Average Pre-Injury Play Time", x = "Pre-Injury Play Time (downs)")

WeekofInjuryUni <- gf_dhistogram(~WeekOfInjury, data = finalData1)%>%
  gf_dens()%>%
  gf_labs(title = "Week of Injury", x = "Week of Injury")

AgeAtConcussionUni <- gf_dhistogram(~AgeAtConcussion, data = finalData1, main = "AgeOfFirstConcussion") %>%
  gf_dens()%>%
  gf_labs(title = "Age at First Concussion", x = "Age at First Concussion")

grid.arrange(PreIPTUni, WeekofInjuryUni, AgeAtConcussionUni, ncol = 3) 
```
```{r, warning = FALSE, echo = FALSE, results='hide'}
favstats(~PreIPT, data = finalData1)
favstats(~WeekOfInjury, data = finalData1)
favstats(~AgeAtConcussion, data = finalData1)
```
**Fig. 3.** The distribution of Pre-Injury Play Time is unimodal and slightly skewed to the left. The median is 51 downs and the IQR is 33.07 downs. The mean is 47.321 downs and the standard deviation is 19.906 downs. Week of Injury looks bimodal and almost uniform, with one main mode at week 10. The median is 9 weeks and the IQR is 7 weeks. The mean is 8.78 weeks and the standard deviation is 4.60 weeks. The distribution of Age at Concussion is unimodal and slightly skewed to the right. The median is 25.88 years and the IQR is 3.91 years. The mean is 26.357 years and the standard deviation is 2.902 years. 


## Bivariate Exploratory Data Analysis

### Qualitative Variables

```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 12}
PositionBiv <- gf_boxplot(PostIPT ~ PositionC, data = finalData1)%>%
  gf_labs(title = "Post-Injury Play-Time vs. Position", x = "Position", y = "Post-Injury Play Time")

AmtgamesmissedBiv <- gf_boxplot(PostIPT ~ amtGamesMissedC, data = finalData1)%>%
  gf_labs(title = "Post-Injury Play-Time vs. Amount of Games Missed", x = "Amount of Games Missed", y = "Post-Injury Play Time")
```
```{r, warning = FALSE, echo = FALSE, fig.show='hide'}
amtgamesmissedmod <- lm(PostIPT ~ amtGamesMissedC, data = finalData1)
amtgamesmissedresids <- mplot(amtgamesmissedmod, which = 1)
amtgamesmissedqq <- mplot(amtgamesmissedmod, which = 2)
```
```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 12}
grid.arrange(PositionBiv, AmtgamesmissedBiv, ncol = 2)
grid.arrange(amtgamesmissedresids, amtgamesmissedqq, ncol = 2)
```
```{r, warning = FALSE, echo = FALSE, results='hide'}
favstats(PostIPT ~ PositionC, data = finalData1)
favstats(PostIPT ~ amtGamesMissedC, data = finalData1)
```
**Fig. 4.** Regarding Position, there does not seem to be a massive difference in the median and IQR of Post-Injury Play Time between offensive players (median = 47, IQR = 39) and defensive players (median = 54, IQR = 44.25). This suggests that a player's position may not play a role in their post-injury play time. However, regarding Amount of Games Missed, the median and IQR of Post-Injury Play Time for players who missed two or more games (median = 23, IQR = 39) looks lower than that of players who missed one game (median = 51, IQR = 43.25) and players who missed no games (median = 49, IQR = 41.25). This suggests that the more games a player misses, the less play time they get after the injury. Since there may be a relationship here, we checked the conditions for Amount of Games Missed: the residuals plot looks good, and the ratio of the largest standard deviation to the smallest standard deviation is less than 2 so we have equal variance. However, the QQ plot reveals that many observations deviate near the top and bottom, so we proceed with extreme caution regarding normality of errors. 

### Quantitative Variables

```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 12}
AgeAtConcussionBiv <- gf_point(PostIPT ~ AgeAtConcussion, data = finalData1)%>%
  gf_lm()%>%
  gf_labs(title = "Post-Injury Play Time vs. Age at Concussion", x = "Age at Concussion", y = "Post-Injury Play Time")

WeekofInjuryBiv <- gf_point(PostIPT ~ WeekOfInjury, data = finalData1)%>%
  gf_lm()%>%
  gf_labs(title = "Post-Injury Play Time vs. Week of Injury", x = "Week of Injury", y = "Post-Injury Play Time")

grid.arrange(AgeAtConcussionBiv, WeekofInjuryBiv, ncol = 2)
```
```{r, warning = FALSE, echo = FALSE, results='hide'}
cor(PostIPT ~ AgeAtConcussion, data = finalData1, use = "pairwise")
cor(PostIPT ~ WeekOfInjury, data = finalData1, use = "pairwise")
```
**Fig. 5.** There is little to no relationship between a player's post-injury play time and the age at which they sustained the concussion (r = 0.0414). There is also little relationship between a player's post-injury play time and the week at which they sustained the concussion (r = 0.00207).

```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 7}
PreIPTOutliers <- gf_point(PostIPT ~ PreIPT, data = finalData1)%>%
  gf_lm()%>%
  gf_labs(title = "Post-Injury Play Time vs. Pre-Injury Play Time (Outliers)", x = "Pre-Injury Play Time", y = "Post-Injury Play Time")
```
```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 12, fig.show='hide'}
#Hid these plots. 
PostIPTvsPreIPTOutliers <- lm(PostIPT ~ PreIPT, data = finalData1)
PostIPTvsPreIPTOutliersResids <- mplot(PostIPTvsPreIPTOutliers, which = 1)
PostIPTvsPreIPTOutliersQQ <- mplot(PostIPTvsPreIPTOutliers, which = 2)
```


```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 15}
grid.arrange(PreIPTOutliers, PostIPTvsPreIPTOutliersResids, PostIPTvsPreIPTOutliersQQ, ncol = 3)
```

**Fig. 6. Pre-Injury Play Time With the Outliers.** After conducting bivariate analysis for Post-Injury Play Time vs. Pre-Injury Play Time, we noticed some unusual points: one at Pre-Injury Play Time = 0 and Post-Injury Play Time > 80, and another at Pre-Injury Play Time at around 83 and Post-Injury Play Time <25. We will evaluate the most extreme outlier and see if conditions improve by only removing one point to preserve numbers. 

```{r warning = FALSE, echo = FALSE, fig.height = 3, fig.width = 4}
plot(PostIPTvsPreIPTOutliers, which = 5)
```

**Fig. 6a. evaluating Pre-Injury Play Time Unusual Points** Observation 30 (Observation 37 in the actual data set) is the most unusual point. This player is Chris Prosinski, identified as an extreme outlier due to his standardized residuals exceeding 4 (threshold is 3) and high leverage over 3.5 (high leverage threshold at 6/203 = 0.0296). Upon researching Prosinski's injury, we found that he played fewer downs before his injury in week 15 and none after. One possible explanation is a data entry error where Prosinski's playtime before and after the injury was reversed. Therefore, we will remove this player and assess if conditions improve.

```{r, warning = FALSE, echo = FALSE}
outliers <-finalData1[c(37), ]
FDfiltered <- finalData1[-c(37),]

PreIPTFilteredPoint<-gf_point(PostIPT ~ PreIPT, data = FDfiltered)%>%
  gf_lm()%>%
  gf_labs(title = "Post-Injury Play Time vs. Pre-Injury Play Time (No Outliers)", x = "Pre-Injury Play Time", y = "Post-Injury Play Time")
```
```{r, warning = FALSE, echo = FALSE, fig.show='hide'}
PreIPTFiltered <- lm(PostIPT ~ PreIPT, data = FDfiltered)
PreIPTFilteredResids <- mplot(PreIPTFiltered, which = 1)
PreIPTFilteredQQ <- mplot(PreIPTFiltered, which = 2)
cor(PostIPT ~ PreIPT, data = FDfiltered, use = "pairwise")
```
```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 18}
grid.arrange(PreIPTFilteredPoint, PreIPTFilteredResids, PreIPTFilteredQQ, ncol = 3)
```
**Fig. 7. Pre-Injury Play Time Without the Outlier.** After removing the outlier, the residuals and QQ plot improved. The residuals plot looks homoscedastic with one point that seems unusual (observation 53). The qq-plot has a slight taper on the lower hand, but looks normally distributed otherwise. There were other points that were part of the taper, so we decided to proceed with caution and keep observation 53. There is a strong, positive, linear relationship between Post-Injury Play Time and Pre-Injury Play Time (r = 0.658). In other words, the more pre-injury play time a player had, the more post-injury play time a player had. 

```{r warning = FALSE, echo = FALSE}
msummary(PreIPTFiltered)
```
**Table 1. SLR Using Pre-Injury Play Time To Predict Post-Injury Play Time** Based on the Model Summary, Pre-Injury Play Time is a significant predictor for Post-Injury Play Time F(1,166) = 170, p-value <0.001. The coefficient of 0.8794 indicates that for every 1 down increase in Pre-Injury Play Time, Post-Injury Play Time will increase on average by 0.8794. Since this number is less than 1, the average play time after an injury will be less than the play time before the injury. 

## Multiple Linear Regression
```{r warning = FALSE, echo = FALSE}
finalData1 <- select(FDfiltered,"PostIPT" , "PreIPT", "WeekOfInjury", "AgeAtConcussion", "PositionC", "amtGamesMissedC" )
KitchenSink <- lm(PostIPT ~  PreIPT + WeekOfInjury + AgeAtConcussion + PositionC + amtGamesMissedC, data = FDfiltered)

car::vif(KitchenSink)

msummary(KitchenSink)
```
**Table 2. Kitchen Sink Model** Based on the VIF<5 for all predictors in the Kitchen Sink Model, there are no issues associated with multicollinearity among predictors. The Kitchen Sink Model is significant at predicting Post-Injury Play Time overall, F(6,161)= 30.3, p-value <0.001. The model summary for the Kitchen Sink indicates that Pre-Injury Play Time (p-value for the t-test <0.0001) and missing two or more games (p-value for t-test = 0.03) are significant predictors for Post-Injury Play Time. Since not all predictors are significant, we will check conditions after reducing the model to two predictors. 

```{r warning = FALSE, echo = FALSE}
ModTwoPred <- lm(PostIPT ~ PreIPT + amtGamesMissedC, data = FDfiltered)
msummary(ModTwoPred)
```
**Table 3. Two Predictor Model With Pre-Injury Play Time and Amount of Games Missed** The MLR model is significant overall, F(3,164) = 60, p-value <0.001. Within the model, Pre-Injury Play Time (p-value <0.001 for the t-test) and Missing two or more games (p-value = 0.015 for the t-test) are significant predictors. Missing one game is not a significant predictor in the MLR model, with a p-value = 0.777 for the t-test. 

```{r  warning = FALSE, echo = FALSE, fig.height=3, fig.width=12}
mod2Resid <- mplot(ModTwoPred, which = 1)
mod2QQ <- mplot(ModTwoPred, which = 2)
grid.arrange(mod2Resid, mod2QQ, ncol = 2)
```

**Fig. 8 Conditions For Two Predictor Model** The residuals plot looks homoskedastic and linear, so the equal variance condition is satisfied. The QQplot has a slight taper on the lower end, but overall looks normal enough. Assuming that the independence condition is satisfied, we can proceed with caution due to the taper on the lower end of the QQplot. 

```{r warning = FALSE, echo = FALSE}
anova(PreIPTFiltered, ModTwoPred, KitchenSink)
```
**Table 4. Nested F-Test** The two predictor model with Pre-Injury Play Time and Amount of Games Missed (categorical) accounts for significantly more variablility in Post-Injury Play Time based on nested f-test, F(2,164) = 3.09 p-value = 0.048. 


# **Interpreting the Final Model**

$\hat{\text{PostIPT}} = 6.6832 + 0.8698(PreIPT) - 0.8171(MissingOneGame) - 10.6248(MissingTwoOrMoreGames)$

Using manual selection from the Kitchen Sink Model (Table 2) and a nested F-test (Table 4), we found that best model used Pre-Injury Play Time and Amount of Games Missed (categorical) as predictors for Post-Injury Play Time (Table 3). We verified that there were no concerns for multicollinarity among predictors with all VIFs <5 (Table 2). Based on Fig. 8, conditions for multiple linear regression are satisfied. Our model accounts for 51.5% of the variability in Post-Injury Play Time and is on average 17.1 downs off of the actual Play Time Post-Injury (Table 3). The coefficient for the Pre-Injury Play Time predictor indicates that for every 1 down increase in Pre-Injury Play Time, Post-Injury Play Time increase on average by 0.8698, after accounting for the Amount of Games Missed. This indicates the Pre-Injury and Post-Injury Play time are positively associated. Although there is a positive relationship between Pre and Post Injury Play Time, the coefficient of 0.8698 indicates that our model predicts player's Post-Injury Play Time to be less than Pre-Injury Play Time overall (since the slope is less than 1). The coefficient for Missing Two Or More Games indicates that players who have missed two or more games will have 10.6248 less downs on average Post-Injury, after accounting for the player's Pre-Injury Play Time. Our final Model is a parallel slopes model that can be separated into three equations based on the amount of games missed. Although only players who have missed two or more games had a significant coefficient, we cannot combine the equation because the significance exists still. 

$\hat{\text{PostIPT(MissingNoGames)}} = 6.6832 + 0.8698(PreIPT))$

$\hat{\text{PostIPT(MissingOneGame)}} = 5.8661 + 0.8698(PreIPT)$

$\hat{\text{PostIPT(MissingTwoOrMoreGames)}} = -3.9416 + 0.8698(PreIPT)$



# **Conclusion**
At the beginning of this study, we sought to identify any variables that may help predict an NFL football player's play time after they recover from their concussion. We ultimately found that  Pre-Injury Play Time and the Amount of Games Missed significantly influenced a player's post-injury play time. More specifically, our final model revealed that the more play time a player had before his concussion, the more play time the player would have after his concussion. The final model additionally revealed that if a player missed two or more games, he would have significantly less post-injury play time than a player who missed no games. 
Our current model only explains 51.5% of the variability in post-injury play time, and consistently under-predicts post-injury play time for players. We believe this may be because our data set was not comprehensive enough and lacked variables that may play a large role in post-concussion play time. For instance, variables like the severity of the concussion could impact post-injury play time; a player with a more severe concussion, despite being cleared to play, may return less healthy than a player with a less severe concussion which might impact their post-injury play time. Additionally, the data set doesn't separately account for players who went to playoffs and therefore played more games than a regular season. Having a variable indicating "yes" - the player had playoff games - and "no" - the player did not have playoff games - could also explain some variability. Additional limitations include the fact that this is an observational study, so we cannot make any causal conclusions, and that the QQ plot for our model is not as normal as we'd like it to be. 
Since we were only barely able to satisfy our conditions, we are considering a Randomization F-test as our next step to see if we can identify any significant relationships in that manner. Another next step is to find a more robust data set, hopefully with more variables and seasons. At the start of the study, we hoped that pre-season injury may help explain some variability in post-injury play time, but we had to remove this variable because of the imbalance of observations for each category. With a larger data set, we might have a more balanced pre-season injury variable, and yet another variable that might help us predict post-injury play time - something that will have significant implications in the world of football. 
