---
title: "Predicting Post Injury Play Time For NFL Players"
author: 'Group D: Arthur, Donna, Kaitlyn, Michael'
date: "5/7/2024"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  word_document: default
---

```{r include=FALSE}
library(mosaic)
library(GGally)
require(tidyr)
require(car)
library(broom)
library(tidyverse)
require(lattice)
library(gridExtra)
library(grid)
library(leaps)
library(lubridate) # finding age at a given date
options(scipen=5)
```

```{r include=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)
trellis.par.set(theme=theme.mosaic())  
options(digits=5)
```


```{r, include=FALSE, warning = FALSE, echo = FALSE}
###############################################################################################
#getwd() in console to find the directory where the csv file is stored on your computer
#For your computer you need to change the path so you can reed the file
#alldata <- read.csv("/home/class26/FIRSTPARTOFEMAIL/ConcussionInjuries2012-2014.csv") 

##Donna's computer
df1 <- read.csv("C:/Users/donna/OneDrive/Documents/STAT230/Stats230FinalProject/ConcussionInjuries2012-2014.csv")
df2 <- read.csv("C:/Users/donna/OneDrive/Documents/STAT230/Stats230FinalProject/HeadInjuredPlayers.csv")


#For Kaitlyn's computer
#df1 <- read.csv("/home/class27/khuang27/Stats230FP/ConcussionInjuries2012-2014.csv")
#df2 <- read.csv("/home/class27/khuang27/Stats230FP/HeadInjuredPlayers.csv")

#For Arthur's computer
#df1 <- read.csv("/home/class26/anwobi26/Stats230_proj/ConcussionInjuries2012-2014.csv")
#df2 <- read.csv("/home/class26/anwobi26/Stats230_proj/HeadInjuredPlayers.csv")

###############################################################################################
##Merge
## gets rid of players who had multiple injuries in the three years
mergeData <- merge(x = df2, y = df1[!duplicated(df1$Player), ], by = "Player", all.x = TRUE)
mergeData <- filter(mergeData, Reported.Injury.Type == "Concussion") ## filter so there's only concussions 

## qual --> quant var
mergeData$Play.Time.After.Injury = parse_number(mergeData$Play.Time.After.Injury)
mergeData$Average.Playtime.Before.Injury = parse_number(mergeData$Average.Playtime.Before.Injury)
###############################################################################################
##rename variables
nameData <- rename(mergeData, OppTeam = Opposing.Team, PostIPT = Play.Time.After.Injury, PreseasonInjury = Pre.Season.Injury., OnWinningTeam = Winning.Team., amtTimeInjured = Weeks.Injured, WeekOfInjury = Week.of.Injury, UnknownInjury = Unknown.Injury., ReportedTypeOfInjury = Reported.Injury.Type, PreIPT = Average.Playtime.Before.Injury, TotalSnaps = Total.Snaps, amtGamesMissed = Games.Missed, GM12 = X2012.2013...Games.Missed, GM13 = X2013.2014...Games.Missed, GM14 = X2014.2015...Games.Missed, I12 = X2012.2013...Number.of.Injuries , I13 = X2013.2014...Number.of.Injuries, I14 = X2014.2015...Number.of.Injuries, TotalInjuries = Total.Number.of.Injuries..2012.2014., PosDuringInjury = Roles.during.injuries, DOB = Date.of.Birth, AgeAtConcussion = Age.first.concussion..2012.2014., totalGamesMissed = Total.Games.Missed..2012.2014., TeamDurConcussion = Team.s..during.concussion.incidents.2012.2014, DOI = Date)

###############################################################################################
## select variables 
studyD <- nameData %>% 
  select(PostIPT, PreseasonInjury, Season, amtGamesMissed, PreIPT, WeekOfInjury, TotalSnaps, Player, DOB, DOI, AgeAtConcussion, Position, Team, OppTeam, OnWinningTeam)

## getting the current age 
# Convert DOB to date objects
studyD$DOB <- as.Date(as.character(studyD$DOB), format = "%d/%m/%Y")

# Calculate current age
given_date <- as.Date("2024-04-01")
current_age_in_years <- as.numeric(interval(studyD$DOB, given_date) / dyears(2))
studyD$CurrentAge <- round(current_age_in_years, 1)

#############################################################################
## this re-categorizes position to offense and defense
studydata <- studyD %>% 
    mutate(PositionC = case_when(
    Position %in% c("Defensive End", "Defensive Tackle", "Safety", "Comerback" ) ~ "defense",
    Position %in% c("Long Snapper", "Wide Receiver", "Tight End", "Linebacker", "Quarterback", "Center", "Guard", "Full Back", "Running Back", "Offensive Tackle") ~ "offense"))
#############################################################################


#############################################################################
## showing why we are removing preseason injury. 
tally(~PreseasonInjury,data = studydata)

m6a <- gf_bar(~PreseasonInjury,data = studydata)%>%
  gf_labs(title = "Fig. 6a. Pre-season Injury", x = "Pre-season Injury (Yes/No)") 

## filters out players who have gotten preseason injury bc not enough observations. 
##population
finalData <- filter(studydata, PreseasonInjury == "No") 

#############################################################################
## data for the season we are predicting
predictData <-filter(studydata, Season == "2014/2015") 

##sample
## selecting years
studydata <- filter(studydata, Season != "2014/2015") ## our sample data
finalData1 <- select(studydata, "PostIPT" ,"PreIPT", "amtGamesMissed", "WeekOfInjury", "AgeAtConcussion", "PreseasonInjury", "Position","Team", "OnWinningTeam", "PositionC", "Player")

# mutates amt games missed to categorical variable. 
finalData1 <- mutate(finalData1, amtGamesMissedC = cut(amtGamesMissed, breaks = c(-1, 0.5, 1.5, 20), labels = c("no games", "one game", "two or more games"), include.lowest = TRUE))

finalData1 <- filter(finalData1, amtGamesMissedC %in% c("no games", "one game", "two or more games"))
```


# **Abstract**
Insert abstract here.


# **Background and Meaning**
Insert stuff here.


# **Methods**

## Data Set
Insert description of data set here.

## Variables
Insert information on variables here.

## Statistical Models
Insert info on statistical models here.


# **Results**

## Univariate Exploratory Data Analysis

### Response Variable

Our response variable for this study is Post-Injury Play Time.  

```{r, warning = FALSE, echo = FALSE, fig.height = 2, fig.width = 3}
gf_dhistogram(~PostIPT, data = finalData1)%>%
  gf_dens()%>%
  gf_labs(title = "Fig. 1. Post-Injury Play Time")
```
```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 12, fig.show='hide'}
favstats(~PostIPT, data = finalData1)
```
**Fig. 1.** The distribution is bimodal, with modes at around 20 downs and 62 downs, and symmetric. The median is 49 downs and the IQR is 42.5 downs. The mean is 47.37 downs and the standard deviation is 24.64 downs.

### Explanatory Variables - Qualitative

Our qualitative explanatory variables are Amount of Games Missed, Position, and Pre-Season Injury.

```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 10}
AmtGamesMissedUni <- gf_bar(~amtGamesMissedC, data = finalData1, fill = c("darkred", "darkblue", "darkgreen")) %>%
  gf_dens() %>%
  gf_labs(title = "Amount of Games Missed (Transformed)",  x = "Amount of Games Missed")

AmtGamesMissedUniQuant <- gf_dhistogram(~amtGamesMissed, data = finalData1)%>%
  gf_dens()%>%
  gf_labs(title = "Amount of Games Missed (Non-Transformed)", x = "Amount of Games Missed")

PositionUni <- gf_bar(~ PositionC, data = finalData1, fill = c("pink", "skyblue"))%>%
  gf_labs(title = "Position", x = "Position")

PreSeasonInjuryUni <- gf_bar(~ PreseasonInjury, data = studydata, fill = c("pink", "darkgreen"))%>%
  gf_labs(title = "Pre-Season Injury", x = "Pre-Season Injury")

grid.arrange(AmtGamesMissedUni, AmtGamesMissedUniQuant, ncol = 2)

grid.arrange(PositionUni, PreSeasonInjuryUni, ncol = 2)
```

```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 12, fig.show='hide'}
favstats(PostIPT ~ amtGamesMissedC, data = finalData1)
favstats(PostIPT ~ PositionC, data = finalData1)
tally(~PreseasonInjury,data = studydata)
```
**Fig. 2** Amount of Games Missed, originally quantitative, is heavily right-skewed, indicating that the majority of players did not miss multiple games. We resultantly transformed it into a categorical variable with the categories "Less than 1 games missed" (54.79%), "One game missed" (33.79%), and "Two or more games missed" (11.42%). Regarding position, the categories and corresponding distribution are: defensive end (5.31%), defensive tackle (3.38%), safety (13.04%), comerback (14.98%), long snapper (0.48%), wide receiver (14.98%), tight end (9.18%), linebacker (10.14%), quarterback (4.35%), center (1.93%), guard (6.76%), full back (0.97%), running back (10.63%) and offensive tackle (3.87%). We transformed this variable into two broader categories: "Offense" (59.82%) and "Defense" (40.18%). Finally, regarding Pre-Season Injury, there was an imbalance of observations; only 9 players sustained a pre-season injury in the first two seasons, while 207 players did not. Thus, we removed this variable for our bivariate analysis. 


### Explanatory Variables - Quantitative

Our quantitative variables are Pre-Injury Play-Time, Week of Injury and Age at Concussion. 

```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 15}
PreIPTUni <- gf_dhistogram(~PreIPT, data = finalData1) %>%
  gf_dens()%>%
  gf_labs(title = "Average Pre-Injury Play Time", x = "Pre-Injury Play Time (downs)")

WeekofInjuryUni <- gf_dhistogram(~WeekOfInjury, data = finalData1)%>%
  gf_dens()%>%
  gf_labs(title = "Week of Injury", x = "Week of Injury")

AgeAtConcussionUni <- gf_dhistogram(~AgeAtConcussion, data = finalData1, main = "AgeOfFirstConcussion") %>%
  gf_dens()%>%
  gf_labs(title = "Age at First Concussion", x = "Age at First Concussion")

grid.arrange(PreIPTUni, WeekofInjuryUni, AgeAtConcussionUni, ncol = 3) 
```
**Fig. 3.** The distribution of Pre-Injury Play Time is unimodal and slightly skewed to the left. The median is 51 downs and the IQR is 32 downs. The mean is 47.418 downs and the standard deviation is 19.577 downs. Week of Injury looks bimodal and almost uniform, with one main mode at week 10. The median is 10 weeks and the IQR is 8 weeks. The mean is 9.48 weeks and the standard deviation is 4.69 weeks. The distribution of Age at Concussion is unimodal and slightly skewed to the right. The median is 26.31 years and the IQR is 4.07 years. The mean is 26.497 years and the standard deviation is 2.8941 years. 


## Bivariate Exploratory Data Analysis

### Qualitative Variables

```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 12}
PositionBiv <- gf_boxplot(PostIPT ~ PositionC, data = finalData1)%>%
  gf_labs(title = "Post-Injury Play-Time vs. Position", x = "Position", y = "Post-Injury Play Time")

AmtgamesmissedBiv <- gf_boxplot(PostIPT ~ amtGamesMissedC, data = finalData1)%>%
  gf_labs(title = "Post-Injury Play-Time vs. Amount of Games Missed", x = "Amount of Games Missed", y = "Post-Injury Play Time")

grid.arrange(PositionBiv, AmtgamesmissedBiv, ncol = 2)
```
**Fig. 4.** Regarding Position, there does not seem to be a massive difference in the median and IQR of Post-Injury Play Time between offensive players (median = 47, IQR = 39) and defensive players (median = 54, IQR = 44.25). This suggests that a player's position may not play a role in their post-injury play time. However, regarding Amount of Games Missed, the median and IQR of Post-Injury Play Time for players who missed two or more games (median = 23, IQR = 39) looks lower than that of players who missed one game (median = 51, IQR = 43. 25) and players who missed no games (median = 49, IQR = 41.25). This suggests that the more games a player misses, the less play time they get after the injury. 

### Quantitative Variables

```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 12}
AgeAtConcussionBiv <- gf_point(PostIPT ~ AgeAtConcussion, data = finalData1)%>%
  gf_lm()%>%
  gf_labs(title = "Post-Injury Play Time vs. Age at Concussion", x = "Age at Concussion", y = "Post-Injury Play Time")

WeekofInjuryBiv <- gf_point(PostIPT ~ WeekOfInjury, data = finalData1)%>%
  gf_lm()%>%
  gf_labs(title = "Post-Injury Play Time vs. Week of Injury", x = "Week of Injury", y = "Post-Injury Play Time")

grid.arrange(AgeAtConcussionBiv, WeekofInjuryBiv, ncol = 2)
```
```{r, warning = FALSE, echo = FALSE, fig.show='hide'}
cor(PostIPT ~ AgeAtConcussion, data = finalData1, use = "pairwise")
cor(PostIPT ~ WeekOfInjury, data = finalData1, use = "pairwise")

```
**Fig. 5.** There is little to no relationship between a player's post-injury play time and the age at which they sustained the concussion (r = 0.0414). There is also little relationship between a player's post-injury play time and the week at which they sustained the concussion (r = 0.00207).

```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 7}
PreIPTOutliers <- gf_point(PostIPT ~ PreIPT, data = finalData1)%>%
  gf_lm()%>%
  gf_labs(title = "Post-Injury Play Time vs. Pre-Injury Play Time (Outliers)", x = "Pre-Injury Play Time", y = "Post-Injury Play Time")
```
```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 12, fig.show='hide'}
#Hid these plots. 
PostIPTvsPreIPTOutliers <- lm(PostIPT ~ PreIPT, data = finalData1)
PostIPTvsPreIPTOutliersResids <- mplot(PostIPTvsPreIPTOutliers, which = 1)
PostIPTvsPreIPTOutliersQQ <- mplot(PostIPTvsPreIPTOutliers, which = 2)

```


```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 15}
grid.arrange(PreIPTOutliers, PostIPTvsPreIPTOutliersResids, PostIPTvsPreIPTOutliersQQ, ncol = 3)
```

**Fig. 6. Pre-Injury Play Time With the Outliers.** After conducting bivariate analysis for Post-Injury Play Time vs. Pre-Injury Play Time, we noticed two unusual points: one at Pre-Injury Play Time = 0 and Post-Injury Play Time > 80, and another at Pre-Injury Play Time at around 83 and Post-Injury Play Time <25. We will evaluate the most extreme outlier and see if conditions improve by only removing one point to preserve numbers. 

```{r warning = FALSE, echo = FALSE, fig.height = 3, fig.width = 4}
plot(PostIPTvsPreIPTOutliers, which = 5)
```

**Fig. 6a. evaluating Pre-Injury Play Time Unusual Points** Observation 30 (Observation 37 in the actual data set) is the most unusual point. This player is Chris Prosinski, identified as an extreme outlier due to his standardized residuals exceeding 4 (threshold is 3) and high leverage over 3.5 (high leverage threshold at 6/203 = 0.0296). Upon researching Prosinski's injury, we found that he played fewer downs before his injury in week 15 and none after. One possible explanation is a data entry error where Prosinski's playtime before and after the injury was reversed. Therefore, we will remove this player and assess if conditions improve.

```{r, warning = FALSE, echo = FALSE}
outliers <-finalData1[c(30), ]
FDfiltered <- finalData1[-c(30),]

PreIPTFilteredPoint<-gf_point(PostIPT ~ PreIPT, data = FDfiltered)%>%
  gf_lm()%>%
  gf_labs(title = "Post-Injury Play Time vs. Pre-Injury Play Time (No Outliers)", x = "Pre-Injury Play Time", y = "Post-Injury Play Time")
```
```{r, warning = FALSE, echo = FALSE, fig.show='hide'}
PreIPTFiltered <- lm(PostIPT ~ PreIPT, data = FDfiltered)
PreIPTFilteredResids <- mplot(PreIPTFiltered, which = 1)
PreIPTFilteredQQ <- mplot(PreIPTFiltered, which = 2)
cor(PostIPT ~ PreIPT, data = FDfiltered, use = "pairwise")
```
```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 18}
grid.arrange(PreIPTFilteredPoint, PreIPTFilteredResids, PreIPTFilteredQQ, ncol = 3)
```
**Fig. 7. Pre-Injury Play Time Without the Outlier.** After removing the outlier, the residuals and QQ plot improved. The residuals plot looks homoscedastic with one point that seems unusual (observation 53). The qq-plot has a slight taper on the lower hand, but looks normally distributed otherwise. There were other points that were part of the taper, so we decided to proceed with caution and keep observation 53. There is a strong, positive, linear relationship between Post-Injury Play Time and Pre-Injury Play Time (r = 0.658). In other words, the more pre-injury play time a player had, the more post-injury play time a player had. 

```{r warning = FALSE, echo = FALSE}
msummary(PreIPTFiltered)
```
**Table 1. SLR Using Pre-Injury Play Time To Predict Post-Injury Play Time** Based on the Model Summary, Pre-Injury Play Time is a significant predictor for Post-Injury Play Time F(1,166) = 170, p-value <0.001. The coefficient of 0.8794 indicates that for every 1 down increase in Pre-Injury Play Time, Post-Injury Play Time will increase on average by 0.8794. Since this number is less than 1, the average play time after an injury will be less than the play time before the injury. 

## Multiple Linear Regression
```{r warning = FALSE, echo = FALSE}
finalData1 <- select(FDfiltered,"PostIPT" , "PreIPT", "WeekOfInjury", "AgeAtConcussion", "PositionC", "amtGamesMissedC" )
KitchenSink <- lm(PostIPT ~  PreIPT + WeekOfInjury + AgeAtConcussion + PositionC + amtGamesMissedC, data = FDfiltered)

car::vif(KitchenSink)

msummary(KitchenSink)
```
**Table 2. Kitchen Sink Model** Based on the VIF<5 for all predictors in the Kitchen Sink Model, there are no issues associated with multicollinearity among predictors. The Kitchen Sink Model is significant at predicting Post-Injury Play Time overall, F(6,161)= 30.3, p-value <0.001. The model summary for the Kitchen Sink indicates that Pre-Injury Play Time (p-value for the t-test <0.0001) and missing two or more games (p-value for t-test = 0.03) are significant predictors for Post-Injury Play Time. Since not all predictors are significant, we will check conditions after reducing the model to two predictors. 

```{r warning = FALSE, echo = FALSE}
ModTwoPred <- lm(PostIPT ~ PreIPT + amtGamesMissedC, data = FDfiltered)
msummary(ModTwoPred)
```
**Table 3. Two Predictor Model With Pre-Injury Play Time and Amount of Games Missed** The MLR model is significant overall, F(3,164) = 60, p-value <0.001. Within the model, Pre-Injury Play Time (p-value <0.001 for the t-test) and Missing two or more games (p-value = 0.015 for the t-test) are significant predictors. Missing one game is not a significant predictor in the MLR model, p-value = 0.777 for the t-test. 

```{r  warning = FALSE, echo = FALSE, fig.height=3, fig.width=12}
mod2Resid <- mplot(ModTwoPred, which = 1)
mod2QQ <- mplot(ModTwoPred, which = 2)
grid.arrange(mod2Resid, mod2QQ, ncol = 2)
```

**Fig. 8 Conditions For Two Predictor Model** The residuals plot looks homoskedastic and linear, so the equal variance condition is satisfied. The QQplot has a slight taper on the lower end, but overall looks normal enough. Assuming that the independence condition is satisfied, we can proceed with caution due to the taper on the lower end of the QQplot. 

```{r warning = FALSE, echo = FALSE}
anova(PreIPTFiltered, ModTwoPred, KitchenSink)
```
**Table 4. Nested F-Test** The two predictor model with Pre-Injury Play Time and Amount of Games Missed (categorical) accounts for significantly more variable in Post-Injury Play Time based on nested f-test, F(2,164) = 3.09 p-value = 0.048. 

## Final Model

# **Interpreting the Final Model**

$\hat{\text{PostIPT}} = 6.6832 + 0.8698(PreIPT) - 0.8171(MissingOneGame) - 10.6248(MissingTwoOrMoreGames)$

Using manual selection from the Kitchen Sink Model (Table 2) and a nested F-test (Table 4), we found that best model used Pre-Injury Play Time and Amount of Games Missed (categorical) as predictors for Post-Injury Play Time (Table 3). We verified that there were no concerns for multicollinarity among predictors with all VIFs <5 (Table 2). Based on Fig. 8, conditions for multiple linear regression are satisfied. Our model accounts for 51.5% of the variability in Post-Injury Play Time and is on average 17.1 downs off of the actual Play Time Post-Injury (Table 3). The coefficient for the Pre-Injury Play Time predictor indicates that for every 1 down increase in Pre-Injury Play Time, Post-Injury Play Time increase on average by 0.8698, after accounting for the Amount of Games Missed. This indicates the Pre-Injury and Post-Injury Play time are positively associated. Although there is a positive relationship between Pre and Post Injury Play Time, the coefficient of 0.8698 indicates that our model predicts player's Post-Injury Play Time to be less than Pre-Injury Play Time overall (since the slope is less than 1). The coefficient for Missing Two Or More Games indicates that players who have missed two or more games will have 10.6248 less downs on average Post-Injury, after accounting for the player's Pre-Injury Play Time. Our final Model is a parallel slopes model that can be separated into three equations based on the amount of games missed. Although only players who have missed two or more games had a significant coefficient, we cannot combine the equation because the significance exists still. 
$\hat{\text{PostIPT(MissingNoGames)}} = 6.6832 + 0.8698(PreIPT))$
$\hat{\text{PostIPT(MissingOneGame)}} = 5.8661 + 0.8698(PreIPT)$
$\hat{\text{PostIPT(MissingTwoOrMoreGames)}} = -3.9416 + 0.8698(PreIPT)$



# **Conclusion**
Put conclusion here.
