---
title: "Season's Information and Post Injury Play Time For NFL Players"
author: 'Group D: Arthur, Donna, Kaitlyn, Michael'
date: "Friday, April 6, 2024"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  word_document: default
---

```{r include=FALSE}
library(mosaic)
library(GGally)
require(tidyr)
library(tidyverse)
require(lattice)
library(gridExtra)
library(grid)
library(leaps)
library(lubridate) # finding age at a given date
options(scipen=5)
```

```{r include=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)
trellis.par.set(theme=theme.mosaic())  
options(digits=5)
```


```{r, include=FALSE, warning = FALSE, echo = FALSE}
###############################################################################################
#getwd() in console to find the directory where the csv file is stored on your computer
#For your computer you need to change the path so you can reed the file
#alldata <- read.csv("/home/class26/FIRSTPARTOFEMAIL/ConcussionInjuries2012-2014.csv") 

##Donna's computer
#df1 <- read.csv("C:/Users/donna/OneDrive/Documents/STAT230/Stats230FinalProject/ConcussionInjuries2012-2014.csv")
#df2 <- read.csv("C:/Users/donna/OneDrive/Documents/STAT230/Stats230FinalProject/HeadInjuredPlayers.csv")


#For Kaitlyn's computer
df1 <- read.csv("/home/class27/khuang27/StatsProjectFinal/ConcussionInjuries2012-2014.csv")
df2 <- read.csv("/home/class27/khuang27/StatsProjectFinal/HeadInjuredPlayers.csv")

###############################################################################################
##Merge
## gets rid of players who had multiple injuries in the three years
mergeData <- merge(x = df2, y = df1[!duplicated(df1$Player), ], by = "Player", all.x = TRUE)
mergeData <- filter(mergeData, Reported.Injury.Type == "Concussion") ## filter so there's only concussions 

## qual --> quant var
mergeData$Play.Time.After.Injury = parse_number(mergeData$Play.Time.After.Injury)
mergeData$Average.Playtime.Before.Injury = parse_number(mergeData$Average.Playtime.Before.Injury)
###############################################################################################
##rename variables
nameData <- rename(mergeData, OppTeam = Opposing.Team, PostIPT = Play.Time.After.Injury, PreseasonInjury = Pre.Season.Injury., OnWinningTeam = Winning.Team., amtTimeInjured = Weeks.Injured, WeekOfInjury = Week.of.Injury, UnknownInjury = Unknown.Injury., ReportedTypeOfInjury = Reported.Injury.Type, PreIPT = Average.Playtime.Before.Injury, TotalSnaps = Total.Snaps, amtGamesMissed = Games.Missed, GM12 = X2012.2013...Games.Missed, GM13 = X2013.2014...Games.Missed, GM14 = X2014.2015...Games.Missed, I12 = X2012.2013...Number.of.Injuries , I13 = X2013.2014...Number.of.Injuries, I14 = X2014.2015...Number.of.Injuries, TotalInjuries = Total.Number.of.Injuries..2012.2014., PosDuringInjury = Roles.during.injuries, DOB = Date.of.Birth, AgeAtConcussion = Age.first.concussion..2012.2014., totalGamesMissed = Total.Games.Missed..2012.2014., TeamDurConcussion = Team.s..during.concussion.incidents.2012.2014, DOI = Date)

###############################################################################################
## select variables 
studyD <- nameData %>% 
  select(PostIPT, PreseasonInjury, Season, amtGamesMissed, PreIPT, WeekOfInjury, TotalSnaps, Player, I12, I13, I14, TotalInjuries, GM12, GM13, GM14,totalGamesMissed, DOB, DOI, AgeAtConcussion, Position)

## getting the current age 
# Convert DOB to date objects
studyD$DOB <- as.Date(as.character(studyD$DOB), format = "%d/%m/%Y")

# Calculate current age
given_date <- as.Date("2024-04-01")
current_age_in_years <- as.numeric(interval(studyD$DOB, given_date) / dyears(2))
studyD$CurrentAge <- round(current_age_in_years, 1)

#############################################################################
## this re-categorizes position to offense and defense
studydata <- studyD %>% 
    mutate(Position = case_when(
    Position %in% c("Defensive End", "Defensive Tackle", "Safety", "Comerback" ) ~ "defense",
    Position %in% c("Long Snapper", "Wide Receiver", "Tight End", "Linebacker", "Quarterback", "Center", "Guard", "Full Back", "Running Back", "Offensive Tackle") ~ "offense"))

#############################################################################
## creating population and sample
predictData <-filter(studydata, Season == "2014/2015") ## data for the season we're predicting
finalData <- filter(studydata, Season != "2014/2015") ## our sample data

finalData1 <- select(finalData, "PostIPT" , "amtGamesMissed", "PreIPT", "WeekOfInjury", "AgeAtConcussion", "PreseasonInjury", "Position")
```

## UNIVARIATE ANALYSIS

### RESPONSE VARIABLE: 
Post-Injury Play Time - Descriptive statistics and a frequency histogram with a density plot are shown below. 
```{r, warning = FALSE, echo = FALSE, fig.height = 2, fig.width = 3}
##PostIPT RESPONSE VARIABLE
favstats(~PostIPT, data = finalData1)
gf_dhistogram(~PostIPT, data = finalData1)%>%
  gf_dens()

```

### EXPLANATORY VARIABLES:
amtGamesMissed, PreIPT, WeekOfInjury, and AgeAtConcussion are shown below as histograms. PreseasonInjury and Position are shown below as tallies.

```{r, warning = FALSE, echo = FALSE}
##EXPLANATORY VARIABLES

## quantitative

m1 <- gf_dhistogram(~amtGamesMissed, data = finalData1)%>%
  gf_dens()

m2 <- gf_dhistogram(~PreIPT, data = finalData1) %>%
  gf_dens()

m3 <- gf_dhistogram(~WeekOfInjury, data = finalData1)%>%
  gf_dens()

m4 <- gf_dhistogram(~AgeAtConcussion, data = finalData1, main = "AgeOfFirstConcussion") %>%
  gf_dens()

##qualitative
m5 <- gf_bar(~ PreseasonInjury, data = finalData1, fill = c("pink", "lightgreen"), main = "Number of Preseason Injuries")

m6 <- gf_bar(~ Position, data = finalData1, fill = c("lavender", "skyblue"), main = "Number of Preseason Injuries")


```
```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 15}
grid.arrange(m1, m2, m3, ncol = 3 ) 
```

```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 15 }
grid.arrange(m4, m5, m6, ncol = 3 ) 
```

## BIVARIATE ANALYSIS

### Post-Injury Play-Time vs. Position. 
We will represent this with a box plot, where the two categories are defense and offense. 

```{r}
gf_boxplot(PostIPT ~ Position, data = finalData1)
```


### Post-Injury Play-Time vs. Pre-Season Injury.
We will represent this with a box plot, in which the two categories are “No,” the player didn’t sustain an injury in the pre-season, and “Yes,” the player did sustain an injury in the pre-season. 

```{r}
gf_boxplot(PostIPT ~ PreseasonInjury, data = finalData1)
```


### Post-Injury Play-Time vs. Pre-Injury Play Time.
We will represent this with a scatterplot. 

```{r}
gf_point(PostIPT ~ PreIPT, data = finalData1)%>%
  gf_lm()
PostIPTvsPreIPT <- lm(PostIPT ~ PreIPT, data = finalData1)
mplot(PostIPTvsPreIPT, which = 1)
mplot(PostIPTvsPreIPT, which = 2)
```

### Post-Injury Play-Time vs. AgeAtConcussion.
We will represent this with a scatterplot.

```{r}
gf_point(PostIPT ~ AgeAtConcussion, data = finalData1)%>%
  gf_lm()
PostIPTvsAge <- lm(PostIPT ~ AgeAtConcussion, data = finalData1)
mplot(PostIPTvsAge, which = 1)
mplot(PostIPTvsAge, which = 2)
```

### Post-Injury Play-Time vs. Week of Injury.
We will represent this with a scatterplot.

```{r}
gf_point(PostIPT ~ WeekOfInjury, data = finalData1)%>%
  gf_lm()
PostIPTvsWeekofInjury <- lm(PostIPT ~ WeekOfInjury, data = finalData1)
mplot(PostIPTvsWeekofInjury, which = 1)
mplot(PostIPTvsWeekofInjury, which = 2)
```

### Post-Injury Play-Time vs. amtGamesMissed.
We will represent this with a scatterplot.

```{r}
gf_point(PostIPT ~ amtGamesMissed, data = finalData1)%>%
  gf_lm()
PostIPTvs.GamesMissed <- lm(PostIPT ~ amtGamesMissed, data = finalData1)
mplot(PostIPTvs.GamesMissed, which = 1)
mplot(PostIPTvs.GamesMissed, which = 2)
```

