---
title: "Season's Information and Post Injury Play Time For NFL Players"
author: 'Group D: Arthur, Donna, Kaitlyn, Michael'
date: "Friday, April 6, 2024"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  word_document: default
---

```{r include=FALSE}
library(mosaic)
library(GGally)
require(tidyr)
require(car)
library(broom)
library(tidyverse)
require(lattice)
#require(kableExtra)
library(gridExtra)
library(grid)
library(leaps)
library(lubridate) # finding age at a given date
options(scipen=5)
```

```{r include=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)
trellis.par.set(theme=theme.mosaic())  
options(digits=5)
```


```{r, include=FALSE, warning = FALSE, echo = FALSE}
###############################################################################################
#getwd() in console to find the directory where the csv file is stored on your computer
#For your computer you need to change the path so you can reed the file
#alldata <- read.csv("/home/class26/FIRSTPARTOFEMAIL/ConcussionInjuries2012-2014.csv") 

##Donna's computer
df1 <- read.csv("C:/Users/donna/OneDrive/Documents/STAT230/Stats230FinalProject/ConcussionInjuries2012-2014.csv")
df2 <- read.csv("C:/Users/donna/OneDrive/Documents/STAT230/Stats230FinalProject/HeadInjuredPlayers.csv")


#For Kaitlyn's computer
#df1 <- read.csv("/home/class27/khuang27/Stats230FP/ConcussionInjuries2012-2014.csv")
#df2 <- read.csv("/home/class27/khuang27/Stats230FP/HeadInjuredPlayers.csv")

#For Arthur's computer
#df1 <- read.csv("/home/class26/anwobi26/Stats230_proj/ConcussionInjuries2012-2014.csv")
#df2 <- read.csv("/home/class26/anwobi26/Stats230_proj/HeadInjuredPlayers.csv")

###############################################################################################
##Merge
## gets rid of players who had multiple injuries in the three years
mergeData <- merge(x = df2, y = df1[!duplicated(df1$Player), ], by = "Player", all.x = TRUE)
mergeData <- filter(mergeData, Reported.Injury.Type == "Concussion") ## filter so there's only concussions 

## qual --> quant var
mergeData$Play.Time.After.Injury = parse_number(mergeData$Play.Time.After.Injury)
mergeData$Average.Playtime.Before.Injury = parse_number(mergeData$Average.Playtime.Before.Injury)
###############################################################################################
##rename variables
nameData <- rename(mergeData, OppTeam = Opposing.Team, PostIPT = Play.Time.After.Injury, PreseasonInjury = Pre.Season.Injury., OnWinningTeam = Winning.Team., amtTimeInjured = Weeks.Injured, WeekOfInjury = Week.of.Injury, UnknownInjury = Unknown.Injury., ReportedTypeOfInjury = Reported.Injury.Type, PreIPT = Average.Playtime.Before.Injury, TotalSnaps = Total.Snaps, amtGamesMissed = Games.Missed, GM12 = X2012.2013...Games.Missed, GM13 = X2013.2014...Games.Missed, GM14 = X2014.2015...Games.Missed, I12 = X2012.2013...Number.of.Injuries , I13 = X2013.2014...Number.of.Injuries, I14 = X2014.2015...Number.of.Injuries, TotalInjuries = Total.Number.of.Injuries..2012.2014., PosDuringInjury = Roles.during.injuries, DOB = Date.of.Birth, AgeAtConcussion = Age.first.concussion..2012.2014., totalGamesMissed = Total.Games.Missed..2012.2014., TeamDurConcussion = Team.s..during.concussion.incidents.2012.2014, DOI = Date)

###############################################################################################
## select variables 
studyD <- nameData %>% 
  select(PostIPT, PreseasonInjury, Season, amtGamesMissed, PreIPT, WeekOfInjury, TotalSnaps, Player, I12, I13, I14, TotalInjuries, GM12, GM13, GM14,totalGamesMissed, DOB, DOI, AgeAtConcussion, Position, Team, OppTeam, OnWinningTeam)

## getting the current age 
# Convert DOB to date objects
studyD$DOB <- as.Date(as.character(studyD$DOB), format = "%d/%m/%Y")

# Calculate current age
given_date <- as.Date("2024-04-01")
current_age_in_years <- as.numeric(interval(studyD$DOB, given_date) / dyears(2))
studyD$CurrentAge <- round(current_age_in_years, 1)

#############################################################################
## this re-categorizes position to offense and defense
studydata <- studyD %>% 
    mutate(PositionC = case_when(
    Position %in% c("Defensive End", "Defensive Tackle", "Safety", "Comerback" ) ~ "defense",
    Position %in% c("Long Snapper", "Wide Receiver", "Tight End", "Linebacker", "Quarterback", "Center", "Guard", "Full Back", "Running Back", "Offensive Tackle") ~ "offense"))

#############################################################################
## creating population and sample
predictData <-filter(studydata, Season == "2014/2015") ## data for the season we're 

#############################################################################

studydata <- filter(studydata, Season != "2014/2015") ## our sample data


```



###Remove variable
```{r}
tally(~PreseasonInjury,data = studydata)

m6a <- gf_bar(~PreseasonInjury,data = studydata)%>%
  gf_labs(title = "Fig. 6a. Pre-season Injury", x = "Pre-season Injury (Yes/No)") 


finalData <- filter(studydata, PreseasonInjury == "No") ## filters out players who have gotten preseason injury bc not enough observations. 


finalData1 <- select(finalData, "PostIPT" ,"PreIPT", "amtGamesMissed", "WeekOfInjury", "AgeAtConcussion", "PreseasonInjury", "Position","Team", "OnWinningTeam", "PositionC", "Player")
```


## UNIVARIATE ANALYSIS


### RESPONSE VARIABLE: 
Post-Injury Play Time - Descriptive statistics and a frequency histogram with a density plot are shown below. 
```{r, warning = FALSE, echo = FALSE, fig.height = 2, fig.width = 3}
##PostIPT RESPONSE VARIABLE
favstats(~PostIPT, data = finalData1)
gf_dhistogram(~PostIPT, data = finalData1)%>%
  gf_dens()%>%
  gf_labs(title = "Fig. 1a. Post-Injury Play Time")

```

### EXPLANATORY VARIABLES:
amtGamesMissed, PreIPT, WeekOfInjury, and AgeAtConcussion are shown below as histograms. PreseasonInjury and Position are shown below as tallies.

```{r, warning = FALSE, echo = FALSE}
##EXPLANATORY VARIABLES

## quantitative

m2a <- gf_dhistogram(~amtGamesMissed, data = finalData1)%>%
  gf_dens()%>%
  gf_labs(title = "Fig. 2a. Amount of Games Missed (Non-Transformed)", x = "Amount of Games Missed")

finalData1 <- mutate(finalData1, amtGamesMissedC = cut(amtGamesMissed, breaks = c(0, 0.5, 1.5, 20), labels = c("less than 1 games", "one game", "two or more games"), include.lowest = TRUE))
m2b <- gf_bar(~amtGamesMissedC, data = finalData1)%>%
  gf_dens()%>%
  gf_labs(title = "Fig. 2b. Amount of Games Missed (Transformed)", x = "Amount of Games Missed")

m3a <- gf_dhistogram(~PreIPT, data = finalData1) %>%
  gf_dens()%>%
  gf_labs(title = "Fig. 3a. Average Pre-Injury Play Time", x = "Pre-Injury Play Time (downs)")

m4a <- gf_dhistogram(~WeekOfInjury, data = finalData1)%>%
  gf_dens()%>%
  gf_labs(title = "Fig. 4a. Week of Injury", x = "Week of Injury")

m5a <- gf_dhistogram(~AgeAtConcussion, data = finalData1, main = "AgeOfFirstConcussion") %>%
  gf_dens()%>%
  gf_labs(title = "Fig. 5a. Age at First Concussion", x = "Age at First Concussion")

##qualitative
m7a <- gf_bar(~ PositionC, data = finalData1, fill = c("lavender", "skyblue"))%>%
  gf_labs(title = "Fig. 7a. Position", x = "Position")




```
```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 15}
grid.arrange(m2a, m2b, m3a, m4a, ncol = 4 ) 
```




```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 15 }
grid.arrange(m5a, m6a, m7a, ncol = 3 ) 
```
```{r}
tally(~Position, data = finalData1 , format = "perc")
tally(~Position, data = finalData1 )
tally(~PositionC, data = finalData1 )
tally(~PositionC, data = finalData1 , format = "perc")

favstats(~WeekOfInjury, data = finalData1)

favstats(~amtGamesMissed, data = finalData1)
favstats(~AgeAtConcussion, data = finalData1)
favstats(~PreIPT, data = finalData1)
```
## Recategorize amount of games missed
```{r}
finalData1 <- mutate(finalData1, amtGamesMissedC = cut(amtGamesMissed, breaks = c(0, 0.5, 1.5, 20), labels = c("less than 1 games", "one game", "two or more games"), include.lowest = TRUE))
```


## BIVARIATE ANALYSIS

### Post-Injury Play-Time vs. PositionC. 
We will represent this with a box plot, where the two categories are defense and offense. 

```{r}
gf_boxplot(PostIPT ~ PositionC, data = finalData1)%>%
  gf_labs(title = "Fig. 1. Post-Injury Play-Time vs. Position")%>%
  gf_labs(x = "Position", y = "Post-Injury Play Time")

gf_dens( ~ PostIPT, color = ~PositionC, data = finalData1)%>%
  gf_labs(title = "Fig. 1. Post-Injury Play-Time vs. Position")
favstats(PostIPT ~ PositionC, data = finalData1)
positionmod <- lm(PostIPT ~ PositionC, data = finalData1)
mplot(positionmod, which = 1)
mplot(positionmod, which = 2)
gf_dens(~ residuals(positionmod))
```




### Post-Injury Play-Time vs. Pre-Injury Play Time.
We will represent this with a scatterplot. 

```{r}
gf_point(PostIPT ~ PreIPT, data = finalData1)%>%
  gf_lm()%>%
  gf_labs(title = "Fig. 3. Post-Injury Play Time vs. Pre-Injury Play Time", x = "Pre-Injury Play Time", y = "Post-Injury Play Time")

PostIPTvsPreIPT <- lm(PostIPT ~ PreIPT, data = finalData1)
mplot(PostIPTvsPreIPT, which = 1)
mplot(PostIPTvsPreIPT, which = 2)
mplot(PostIPTvsPreIPT, which = 5)
cor(PostIPT ~ PreIPT, data = finalData1, use = "pairwise")

## residuals looks pretty good other than 2 outlier, qq looks also good enigh but the same 2 outliers at the end
## 25 (Chris Prosinski) has high standardized residuals (over 4) 54 (Erin Henderson) has high standarized (-2.5ish) residuals

outliers <-finalData1[c(30,69), ]
FDfiltered <- finalData1[-c(30,69), ]


gf_point(PostIPT ~ PreIPT, data = FDfiltered)%>%
  gf_lm()%>%
  gf_labs(title = "Fig. 3. Post-Injury Play Time vs. Pre-Injury Play Time", x = "Pre-Injury Play Time", y = "Post-Injury Play Time")

PreIPTFiltered <- lm(PostIPT ~ PreIPT, data = FDfiltered)
mplot(PreIPTFiltered, which = 1)
mplot(PreIPTFiltered, which = 2)
gf_dens(~ residuals(PreIPTFiltered))
```

### Post-Injury Play-Time vs. AgeAtConcussion.
We will represent this with a scatterplot.

```{r}
gf_point(PostIPT ~ AgeAtConcussion, data = finalData1)%>%
  gf_lm()%>%
  gf_labs(title = "Fig. 4. Post-Injury Play Time vs. Age at Concussion", x = "Age at Concussion", y = "Post-Injury Play Time")
PostIPTvsAge <- lm(PostIPT ~ AgeAtConcussion, data = finalData1)
mplot(PostIPTvsAge, which = 1)
mplot(PostIPTvsAge, which = 2)
cor(PostIPT ~ AgeAtConcussion, data = finalData1, use = "pairwise")
```

```{r}
gf_point(PostIPT ~ AgeAtConcussion, data = finalData1)%>%
  gf_lm()%>%
  gf_labs(title = "Post-Injury Play Time vs. Age at Concussion")
PostIPTvsAge <- lm(PostIPT ~ AgeAtConcussion, data = finalData1)
mplot(PostIPTvsAge, which = 1)
mplot(PostIPTvsAge, which = 2)
msummary(PostIPTvsAge)
cor(PostIPT ~ AgeAtConcussion, data = finalData1, use = "pairwise")
```



### Post-Injury Play-Time vs. Week of Injury.
We will represent this with a scatterplot.

```{r}
gf_point(PostIPT ~ WeekOfInjury, data = finalData1)%>%
  gf_lm()%>%
  gf_labs(title = "Fig. 5. Post-Injury Play Time vs. Week of Injury", x = "Week of Injury", y = "Post-Injury Play Time")
PostIPTvsWeekofInjury <- lm(PostIPT ~ WeekOfInjury, data = finalData1)
mplot(PostIPTvsWeekofInjury, which = 1)
mplot(PostIPTvsWeekofInjury, which = 2)
cor(PostIPT ~ WeekOfInjury, data = finalData1, use = "pairwise")
```

### Post-Injury Play-Time vs. amtGamesMissed.
We will represent this with a scatterplot.

```{r}
gf_point(PostIPT ~ amtGamesMissed, data = finalData1)%>%
  gf_lm()%>%
  gf_labs(title = "Post-Injury Play Time vs. Amount of Games Missed")

PostIPTvs.GamesMissed <- lm(PostIPT ~ amtGamesMissed, data = finalData1)
mplot(PostIPTvs.GamesMissed, which = 1)
mplot(PostIPTvs.GamesMissed, which = 2)
cor(PostIPT ~ amtGamesMissed, data = finalData1, use = "pairwise")
```
### Post-Injury Play-Time vs. amtGamesMissedC.
We will represent this with a boxplot
```{r}
gf_boxplot(PostIPT ~ amtGamesMissedC, data = finalData1)%>%
  gf_labs(title = "Fig. 2. Post-Injury Play-Time vs. Amount of Games Missed")%>%
  gf_labs(x = "Amount of Games Missed", y = "Post-Injury Play Time")

gf_dens( ~ PostIPT, color = ~amtGamesMissedC, data = finalData1)%>%
  gf_labs(title = "Post-Injury Play-Time vs. amtGamesMissed")
favstats(PostIPT ~ amtGamesMissedC, data = finalData1)
amtgamesmissedmod <- lm(PostIPT ~ amtGamesMissedC, data = finalData1)
mplot(amtgamesmissedmod, which = 2)
gf_dens(~ residuals(amtgamesmissedmod))
```


##Final Model (As of now)


```{r}

#ggpairs(finalData1)
finalData1 <- select(finalData1,"PostIPT" , "PreIPT", "WeekOfInjury", "AgeAtConcussion", "PositionC", "amtGamesMissed", "amtGamesMissedC" )


fm.full <- lm(PostIPT ~ PreIPT  + WeekOfInjury + AgeAtConcussion + PositionC + amtGamesMissedC, data = finalData1)
anova(fm.full)

best <- regsubsets(PostIPT ~ PreIPT  + amtGamesMissedC + WeekOfInjury + AgeAtConcussion + PositionC ,data = finalData1, nbest = 1)
with(summary(best), data.frame(rsq, adjr2, cp, outmat))


modA <- lm(PostIPT ~ PreIPT + amtGamesMissedC + PositionC, data = finalData1)
anova(modA)
msummary(modA)
mplot(modA, which = 1 )
mplot(modA, which = 2 )
```

```{r, echo = FALSE, fig.height = 4, fig.width = 3}

mod1 <- lm(PostIPT ~ PreIPT + amtGamesMissedC, data = finalData1)
anova(mod1)
msummary(mod1)
mplot(mod1, which = 1 :2 )


##2 outleirs - 25, 54
```
```{r}



FDfiltered <- finalData1[-c(25, 54), ]

```


```{r}
full  <- lm(PostIPT ~ PreIPT + amtGamesMissedC +WeekOfInjury + AgeAtConcussion + PositionC , data = FDfiltered)
anova(full)
three  <- lm(PostIPT ~ PreIPT + amtGamesMissedC + PositionC , data = FDfiltered)
twoPred  <- lm(PostIPT ~ PreIPT + amtGamesMissedC , data = FDfiltered)

onePred  <- lm(PostIPT ~ PreIPT  , data = FDfiltered)

anova(onePred, twoPred, three, full)

anova(twoPred)
mplot(twoPred, which = 1 )
mplot(twoPred, which = 2 )
mplot(twoPred, which = 4)
mplot(twoPred, which = 5 )
mplot(twoPred, which = 6 )
```

## Final Model: 2 predictors, with PreIPT and amtGamesMissed
```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 3}
anova(twoPred)
msummary(twoPred)
mplot(twoPred, which = 1:2 )

mplot(twoPred, which = 4)

#car::vif(fm.full)
```
MLR:

First we ran the kitchen sink model and fond 2 predictors (PreIPT(p-value for t-test = <2e-16) and amtGamesMissed(p-value for t-test = 0.014)) to be significant. We ran VIF on the kitchen sink to see if there were any issues with multicolinarity between our predictions, and all of the VIF < 5, so we had no issues with multicolinarity between our predictors. We then ran a 2 predictor model with only the significant predictors (PreIPT and amtGamesMissed). For equal variance, there were 2 observations with high residuals that we decided to remove - Cecil Shorts and Devin Hester. Cecil Shorts had injuries from the past that he might have been recovering from, which might be why he had higher post injury play time than pre injury play time. Devin Hester is considered to be the greatest return specialist of all time, which might be why he also had significantly more play time post injury than pre injury. 

After removing the two outliers from the dataset, we ran a nested f-test to see if the two predictor model was significantly better than the one predictor model. Position was close to being a significant predictor(p-value = 0.086), so we ran a nested f-test with one(PreIPT), two(PreIPT and amtGamesMissed), three predictors(PreIPT,amtGamesMissed, and position) and the full model(PreIPT,amtGamesMissed, position, WeekOfInjury, and AgeAtConcussion) to see how many variables are worth adding. For the nested f-test, only the two predictor model was significant (p-value = 0.012) so we decided to go with the two predictor model with outliers removed for our final model. 

Conditions: 

For equal variance, this condition looks good enough with no apparent pattern and one unusual point. We used cook's distance to how unusual the point is and if we should remove it. No cook's distance was above 0.5 to be considered an moderately influential, so we decided not to remove the point. For normality, the qq plot looks good enough, with most of the points on the theoretical line and the ends tapering off. 

Final Model (as of now):

Our current model is PostIPT(hat) = 9.4480 + 0.8325(PreIPT) - 3.5095(amtGamesMissed). The Rsquared is 47.6% and the standard error is 17.9. 

```{r}
#Mike Mitchell
pred <- makeFun(twoPred)

pred(PreIPT = 32, amtGamesMissedC = "less than 1 games")
pred(PreIPT = 32, amtGamesMissedC = "less than 1 games", interval = "prediction", level = 0.95)

## last slide show example of player and show how the model did

#Julian Edelman
pred <- makeFun(twoPred)

pred(PreIPT = 41, amtGamesMissedC = "less than 1 games")
pred(PreIPT = 41, amtGamesMissedC = "less than 1 games", interval = "prediction", level = 0.95)
```

