---
title: "Season's Informaiton and Post Injury Play Time For NFL Players"
author: 'Group D: Arthur, Donna, Kaitlyn, Michael'
date: "Firday, March 16, 2024"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  word_document: default
---

```{r include=FALSE}
library(mosaic)
require(tidyr)
library(tidyverse)
require(lattice)
library(gridExtra)
library(grid)
library(leaps)
options(scipen=5)
```

```{r include=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).
# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)
trellis.par.set(theme=theme.mosaic())  
options(digits=5)
```



## Proposal    
**PURPOSE**

Concussions have been a trending topic in football for the last few years as current and former players suffer the consequences of such a physical sport. The NFL has been aware of this and have taken measures such as implementing a concussion protocol that players must follow and complete before returning to play. There has also been more investment in helmet technology to protect players from serious head injuries. But due to the physical nature of the sport concussions are still common today. From a players perspective after sustaining a concussion you want to get back on the field as soon as possible, while also playing at the same level you were playing at before the injury. Additionally, the NFL has made some changes to the rules and structure of the game over the years that may or may not affect the frequency of concussions and how a player performs once they return. We hope that using this data set could help players understand how to maximize their playtime time post injury, and also see if there is enough evidence to make an argument for shortening the NFL season, getting rid of pre-season games, or another change that could help maximize post injury playtime for players.
 
**DATA**

We will be using data from Concussions in the NFL (2012-2014) which contains a dataset concerning NFL players who sustained concussions, head injuries, and illness during the season from the years 2012 to 2014. This dataset details a number of factors surrounding these players and their injuries such as pre-season injury history, position, average play time after injury, the number of weeks injured, the season they got injured, and more. We will filter the dataset for only concussed players to determine the impact of concussions on players' total playing time. 

**POPULATION** 

The population of this dataset is NFL players during all three seasons (2012-2014) that obtained a concussion. Our sample will consist of concussion data from the first two seasons, which we will use to predict the third season.

**RESPONSE VARIABLE**

We chose Post Injury Play Time (PostIPT) as our response variable. This is a quantitative variable, and is represented by the average number of downs per game after the player returns from the injury. The distribution is bi-modal; many observations are close to 0 downs - likely influenced by players who did not return from their injury. The median is 51 downs and the IQR is 40 downs.
 
**EXPLANATORY VARIABLES**

**Pre-season injury (PreseasonInjury)** is a qualitative variable, categorized by Yes – the player sustained an injury before the season began, and No – the player did not sustain an injury before the season. There are only 11 preseason injuries and 242 injuries that did not happen during the preseason, so we might need to filter the data.

**Player position (Position)** is a qualitative variable, categorized by all the possible positions on a football team. We categorized the positions to defense and offense. There are 95 defensive injuries and 158 offensive injuries. 

**Season** is a qualitative variable, categorized in (start of season/end of season) format: 2012/2013 and 2013/2014. There are more injuries from the 2013/2014 season. There are 105 injuries from 2012/2013, and 148 injuries from 2013/2014. 

**Number of games missed (amtGamesMissed)** is a quantitative variable, measured by how many games in one season a player missed after sustaining the injury. This data is skewed to the right, with a median of 1 game missed and an IQR of 1 game missed. We need to transform this data for analysis, possibly taking the square root. 

**Average play time (PreIPT)** before the injury is a quantitative variable, measured in the average number of downs per game that occurred before the player sustained the injury. The distribution is more normally distributed than Post Injury play time and slightly skewd to the left, the median is 51 downs and the IQR is 40 downs. There are no unusual points. 
 
**Week of injury (WeekOfInjury)** is a quantitative variable, measured in the number of weeks into the football season that the player sustained the injury. The data seems to be bimodal, with a median of week 10 when the injury occurred and an IQR of week 8 when the injury occurred. 


```{r, warning = FALSE, echo = FALSE, include=FALSE}
#getwd() in console to find the directory where the csv file is stored on your computer
#For your computer you need to change the path so you can reed the file
#alldata <- read.csv("/home/class26/FIRSTPARTOFEMAIL/ConcussionInjuries2012-2014.csv") 

#For Donna's computer
alldata <- read.csv("C:/Users/donna/OneDrive/Documents/STAT230/Stats230FinalProject/ConcussionInjuries2012-2014.csv") 

#For Kaitlyn's computer
#alldata <- read.csv("/home/class27/khuang27/StatsProjectFinal/ConcussionInjuries2012-2014.csv")

glimpse(alldata)
names(alldata)
```

```{r, warning = FALSE, echo = FALSE, include=FALSE}

alldata <- filter(alldata, Reported.Injury.Type == "Concussion") ## filter so there's only concussions 
predictdata <-filter(alldata, Season == "2014/2015") ## data for the season we're predicting
alldata <- filter(alldata, Season != "2014/2015") ## our sample data

studyD <- rename(alldata, OppTeam = Opposing.Team, PostInjuryPlayTime = Play.Time.After.Injury, PreseasonInjury = Pre.Season.Injury., OnWinningTeam = Winning.Team., amtTimeInjured = Weeks.Injured, WeekOfInjury = Week.of.Injury, UnknownInjury = Unknown.Injury., ReportedTypeOfInjury = Reported.Injury.Type, AveragePreInjuryPlayTime = Average.Playtime.Before.Injury, TotalSnaps = Total.Snaps, amtGamesMissed = Games.Missed)
studyD <- studyD %>% 
  select(PostInjuryPlayTime, PreseasonInjury, Position, Season, amtGamesMissed, AveragePreInjuryPlayTime, WeekOfInjury )

studyD$PostIPT = parse_number(studyD$PostInjuryPlayTime)

studyD$PreIPT = parse_number(studyD$AveragePreInjuryPlayTime)

```

```{r,warning = FALSE, echo = FALSE, include=FALSE}
## this re-categorizes position to offense and defense
studydata <- studyD %>% 
    mutate(Position = case_when(
    Position %in% c("Defensive End", "Defensive Tackle", "Safety", "Comerback" ) ~ "defense",
    Position %in% c("Long Snapper", "Wide Receiver", "Tight End", "Linebacker", "Quarterback", "Center", "Guard", "Full Back", "Running Back", "Offensive Tackle") ~ "offense"))

```



## Exploratory Data Analysis 

```{r, warning = FALSE, echo = FALSE, fig.height = 4, fig.width = 15}
m1 <- gf_histogram(~WeekOfInjury, data = studydata, main = "Distribution of Week of Injury")

m3 <- gf_bar(~ PreseasonInjury, data = studydata, fill = c("darkred", "forestgreen"), main = "Number of Preseason Injuries")

m4 <- gf_bar(~Season, data = studydata)

m5 <- gf_histogram(~amtGamesMissed, data = studydata)

grid.arrange(m1, m3, m4, m5, ncol = 4 ,heights = .5) 

m6 <- gf_dhistogram(~PreIPT, data = studydata) %>%
  gf_dens()
m7 <- gf_dhistogram(~PostIPT, data = studydata)%>%
  gf_dens()

grid.arrange(m6,m7, ncol = 2 ,heights = .5)

```
```{r}
tally(~ Position, data = studydata)
tally(~ PreseasonInjury, data = studydata)
tally(~ Season, data = studydata)

favstats(~WeekOfInjury, data = studydata) # summary stats for the week of the injury

favstats(~amtGamesMissed, data = studydata, title = "Games Missed") # summary stats for games missed

favstats(~PreIPT, data = studydata, main = "Pre Injury Play Time") # summary stats for pre injury play time
favstats(~PostIPT, data = studydata, main = "Post Injury Play Time") # summary stats for post injury play time

#gf_boxplot(~PreIPT, data = studydata)
#gf_boxplot(~PostIPT, data = studydata)
```

